---
title: "Creating Your First Oncoplot"
format: html
editor: visual
---

# Creating your first oncoplot

## Step 1: Install Required Packages 

Only have to run this once.

Tip: run one line at a time and make sure install works correctly.

```{r, eval=FALSE}
install.packages("tidyverse")
install.packages("remotes")
install.packages("readxl")
install.packages("somaticflags")
install.packages("ggiraph")
install.packages("patchwork")
install.packages('ggEDA', repos = c('https://ccicb.r-universe.dev'))
install.packages('ggoncoplot', repos = c('https://selkamand.r-universe.dev'))
```

## Step 2: Load Libraries

Libraries are 'premade' pieces of other peoples code.

```{r}
# For reading excel files
library(readxl)

# Miscellaneous EDA
library(tidyverse)

# Making our Oncoplot
library(ggoncoplot)

# Exclude Noisy Genes
library(somaticflags)

# Create relative filepaths
library(here)
```

## Step 3: Read Data

We'll be using a glioblastoma TCGA dataset (`data/gbm.xlsx`)

```{r}
# Read Genomic Data
gbm <- read_excel(here("data/gbm.xlsx"))

# Print First few rows
head(gbm)

# See the whole dataset (uncomment and run)
View(gbm)
```

## Step 4: Create Oncoplot

### Basic Oncoplot

```{r}
ggoncoplot(gbm, col_genes = "Gene", col_samples = "Sample")
```

### Show all samples

By default many oncoplots will only show unaltered samples. To show every sample in your cohort set `show_all_samples = TRUE`

```{r}
ggoncoplot(
  gbm, 
  col_genes = "Gene", 
  col_samples = "Sample",
  show_all_samples = TRUE
)
```

### Colour based on mutation type

```{r}
ggoncoplot(
  gbm, 
  col_genes = "Gene", 
  col_samples = "Sample", 
  col_mutation_type = "Variant_Classification", # Colours based on mutation type
)
```

### Customise tooltips

```{r}
ggoncoplot(
  gbm, 
  col_genes = "Gene", 
  col_samples = "Sample", 
  col_mutation_type = "Variant_Classification", # Colours based on mutation type
  col_tooltip = "Mutation" # Add mutations to tooltip
)
```

### Add marginal plots

```{r}
ggoncoplot(
  gbm, 
  col_genes = "Gene", 
  col_samples = "Sample", 
  col_mutation_type = "Variant_Classification", 
  draw_gene_barplot = TRUE,
  draw_tmb_barplot = TRUE,
)
```

### Add clinical metadata

We'll read in our data from `data/gbm_clinical.xlsx`

```{r}
# Read Genomic Data
clinical <- read_excel(here("data/gbm_clinical.xlsx"))

# List all columns
colnames(clinical)

# Print First few rows
head(clinical)

# See the whole dataset (uncomment and run)
# View(clinical)
```

As long as it contains a column name that matches the value of `col_samples`, we can just supply entire dataset to the \`metadata\` argument and choose which columns to plot

```{r}
ggoncoplot(
  gbm, 
  col_genes = "Gene", 
  col_samples = "Sample", 
  col_mutation_type = "Variant_Classification",
  metadata = clinical, 
  cols_to_plot_metadata = c("histological_type", "gender", "vital_status")
)
```

Custom palettes and sorting by metadata

```{r}
ggoncoplot(
  gbm, 
  col_genes = "Gene", 
  col_samples = "Sample", 
  col_mutation_type = "Variant_Classification",
  draw_tmb_barplot = TRUE,
  metadata = clinical, 
  cols_to_plot_metadata = c("histological_type", "gender", "vital_status"),
  metadata_palette = list(
    "gender" = c("MALE" = "blue", "FEMALE" = "pink"),
    "vital_status" = c("Alive" = "green", "Dead" = "maroon")
    ), 
   metadata_sort_cols = c("gender")
)
```

## Step 4: Multiomics!!

See [manual](https://selkamand.github.io/ggoncoplot/articles/manual.html) for more info.

Will need to install express first with

```         
install.packages("remotes")
remotes::install_github("selkamand/express")
```

```{r}
library(express)
library(patchwork)
library(ggiraph)

oncoplot <- ggoncoplot(
  gbm, 
  col_genes = "Gene", 
  col_samples = "Sample", 
  col_mutation_type = "Variant_Classification",
  metadata = clinical, 
  draw_tmb_barplot = TRUE,
  cols_to_plot_metadata = c("histological_type", "gender", "vital_status"),
  interactive = FALSE
)

tsne_expression <- express_precomputed("GBM", datatype = "expression", interactive = FALSE)

## Combine plots with patchwork
combined_plots <- (tsne_expression / oncoplot)  + plot_layout(heights = c(2, 8))

# View the interactive version with ggiraph
interactive_multiplot <- girafe(ggobj = combined_plots, height_svg = 6, width_svg = 10)

# Add some settings to choose how to make combined plots interactive
interactive_multiplot <- girafe_options(x = interactive_multiplot,
    opts_selection(type = "multiple", only_shiny = FALSE, css = "opacity: 1"),
    opts_selection_inv(css = "opacity: 0.12")
    )

interactive_multiplot
```
